---
title: "FG Project"
author: "Mitch Jeter"
date: "2025-02-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}

library(nflfastR)
library(dplyr)

#play-by-play data for 2022–2024
pbp_data <- load_pbp(seasons = 2022:2024)

# Filter for field goal attempts
fg_data <- pbp_data %>%
  filter(play_type == "field_goal")

# Add high-leverage situations (e.g., 4th quarter, close games, overtime)
fg_data <- fg_data %>%
  mutate(clutch = ifelse(
    (game_seconds_remaining <= 120 | game_half == "Overtime") & abs(score_differential) <= 3,
    TRUE,
    FALSE
  ))



```

```{r}
write.csv(fg_data, "fg_data.csv")
```


NFL Fast R

Field Goal Performance 
Field goal results and attempts:

field_goal_result
kick_distance
field_goal_attempt

Kicker-specific data:
kicker_player_id
kicker_player_name

Success probabilities:
fg_prob (field goal probability)
vegas_fg_prob

Game outcomes related to field goals:

wpa (Win Probability Added)
vegas_wpa



Game Time and Context
Time-related stats:
quarter_seconds_remaining
half_seconds_remaining
game_seconds_remaining

Timeouts and pressure moments:
home_timeouts_remaining
away_timeouts_remaining
timeout_team
quarter_end

Score differential and team performance:
score_differential
score_differential_post

Win probabilities and adjustments:
wp (Win Probability)
home_wp (Home team Win Probability)
away_wp (Away team Win Probability)

```{r}
# Filter for clutch kicks
team_clutch_stats <- fg_data %>%
  filter(clutch == TRUE) %>%
  group_by(posteam) %>%
  summarize(
    total_clutch_kicks = n(),
    clutch_makes = sum(field_goal_result == "made"),
    clutch_fg_pct = mean(field_goal_result == "made"),
    total_wpa = sum(wpa, na.rm = TRUE)
  ) %>%
  mutate(
    clutch_score = clutch_fg_pct * 0.5 + (total_wpa / total_clutch_kicks) * 0.5
  ) %>%
  arrange(desc(clutch_score))

# Add ranks
team_clutch_stats <- team_clutch_stats %>%
  mutate(rank = row_number())

# View results
print(team_clutch_stats)

```
```{r}
# Filter teams with 100% FG percentage for clutch kicks
teams_100_percent_clutch <- team_clutch_stats %>%
  filter(clutch_fg_pct == 1)

# View results
print(teams_100_percent_clutch)

```




```{r}
# Load libraries
library(tidyverse)
library(ggimage)

# Team abbreviations and their corresponding logo URLs
team_logos <- data.frame(
  posteam = c(
    "ARI", "ATL", "BAL", "BUF", "CAR", "CHI", "CIN", "CLE",
    "DAL", "DEN", "DET", "GB", "HOU", "IND", "JAX", "KC",
    "LAC", "LAR", "LV", "MIA", "MIN", "NE", "NO", "NYG",
    "NYJ", "PHI", "PIT", "SF", "SEA", "TB", "TEN", "WAS"
  ),
  logo_url = c(
    "https://a.espncdn.com/i/teamlogos/nfl/500/ari.png",
    "https://a.espncdn.com/i/teamlogos/nfl/500/atl.png",
    "https://a.espncdn.com/i/teamlogos/nfl/500/bal.png",
    "https://a.espncdn.com/i/teamlogos/nfl/500/buf.png",
    "https://a.espncdn.com/i/teamlogos/nfl/500/car.png",
    "https://a.espncdn.com/i/teamlogos/nfl/500/chi.png",
    "https://a.espncdn.com/i/teamlogos/nfl/500/cin.png",
    "https://a.espncdn.com/i/teamlogos/nfl/500/cle.png",
    "https://a.espncdn.com/i/teamlogos/nfl/500/dal.png",
    "https://a.espncdn.com/i/teamlogos/nfl/500/den.png",
    "https://a.espncdn.com/i/teamlogos/nfl/500/det.png",
    "https://a.espncdn.com/i/teamlogos/nfl/500/gb.png",
    "https://a.espncdn.com/i/teamlogos/nfl/500/hou.png",
    "https://a.espncdn.com/i/teamlogos/nfl/500/ind.png",
    "https://a.espncdn.com/i/teamlogos/nfl/500/jax.png",
    "https://a.espncdn.com/i/teamlogos/nfl/500/kc.png",
    "https://a.espncdn.com/i/teamlogos/nfl/500/lac.png",
    "https://a.espncdn.com/i/teamlogos/nfl/500/lar.png",
    "https://a.espncdn.com/i/teamlogos/nfl/500/lv.png",
    "https://a.espncdn.com/i/teamlogos/nfl/500/mia.png",
    "https://a.espncdn.com/i/teamlogos/nfl/500/min.png",
    "https://a.espncdn.com/i/teamlogos/nfl/500/ne.png",
    "https://a.espncdn.com/i/teamlogos/nfl/500/no.png",
    "https://a.espncdn.com/i/teamlogos/nfl/500/nyg.png",
    "https://a.espncdn.com/i/teamlogos/nfl/500/nyj.png",
    "https://a.espncdn.com/i/teamlogos/nfl/500/phi.png",
    "https://a.espncdn.com/i/teamlogos/nfl/500/pit.png",
    "https://a.espncdn.com/i/teamlogos/nfl/500/sf.png",
    "https://a.espncdn.com/i/teamlogos/nfl/500/sea.png",
    "https://a.espncdn.com/i/teamlogos/nfl/500/tb.png",
    "https://a.espncdn.com/i/teamlogos/nfl/500/ten.png",
    "https://a.espncdn.com/i/teamlogos/nfl/500/was.png"
  )
)

# Assuming team_clutch_stats is your data with clutch scores
team_clutch_stats <- team_clutch_stats %>%
  left_join(team_logos, by = "posteam")

clutch_plot <- team_clutch_stats %>%
  ggplot(aes(x = reorder(posteam, clutch_score), y = clutch_score)) +
  geom_image(aes(image = logo_url), size = 0.1, by = "width") +  # Only show the logos
  coord_flip() +
  labs(
    title = "NFL Team Clutch Kicker Rankings (2022-2024)",
    subtitle = "Based on Clutch FG% and WPA in High-Leverage Situations",
    x = "Team",
    y = "Clutch Score"
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 12),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_blank(),  
    plot.title = element_text(face = "bold", size = 16)
  )

# Print the plot
print(clutch_plot)

```

```{r}
# Assuming team_clutch_stats is already calculated and includes logos

# Rank teams by the most clutch kicks
team_clutch_stats <- team_clutch_stats %>%
  arrange(desc(total_clutch_kicks)) %>%
  mutate(rank = row_number())

# Plot the teams by the most clutch kicks with logos
clutch_kicks_plot <- team_clutch_stats %>%
  ggplot(aes(x = reorder(posteam, total_clutch_kicks), y = total_clutch_kicks)) +
  geom_image(aes(image = logo_url), size = 0.1, by = "width") +  # Only show the logos
  coord_flip() +
  labs(
    title = "NFL Teams Ranked by Total Clutch Kicks (2022-2024)",
    subtitle = "Based on the Number of Clutch Field Goals Attempted",
    x = "Team",
    y = "Total Clutch Kicks"
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 12),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_blank(),  # Hide the y-axis text
    plot.title = element_text(face = "bold", size = 16)
  )

# Print the plot
print(clutch_kicks_plot)

```





MODEL

How to determine what factors create missed FGS 
Is your team more liklet to miss a FG in the clutch

```{r}
kick_made <- as.numeric(as.factor(fg_data$field_goal_result))
```


```{r}
table(fg_data$field_goal_result, kick_made)
```

```{r}
fg_data <- fg_data[which(fg_data$field_goal_result != "blocked"),]
```

```{r}
kick_made <- abs(as.numeric(as.factor(fg_data$field_goal_result)) - 2)
```

```{r}
fg_data$wind[is.na(fg_data$wind)] <- 0 
```


```{r}
table(fg_data$field_goal_result, kick_made)
```





```{r}
model_data <- fg_data[, c("kick_distance", "wind", "quarter_seconds_remaining",
"half_seconds_remaining","score_differential",
"wp", 
"home_wp",  "away_wp", 
"game_seconds_remaining", "clutch")]
```


```{r}
library(xgboost)
```

```{r}
dtrain <- xgb.DMatrix(data = as.matrix(model_data), label = kick_made)
```


```{r}
xgb_1 <- xgboost(dtrain, 
                 objective = "binary:logistic", 
                 nrounds = 100)
```

```{r}
# Extract importance
imp_mat <- xgb.importance(model = xgb_1)
# Plot importance (top 10 variables)
xgb.plot.importance(imp_mat, top_n = 10)
```


```{r}
#install.packages("splitstackshape")
#install.packages("SHAPforxgboost")
source("~/Downloads/a_insights_shap_functions.r", echo=TRUE)

library(splitstackshape)
library(ggforce)
library(SHAPforxgboost)

#Calculate SHAP importance
shap_result_1 <- shap.score.rank(xgb_model = xgb_1,
X_train = as.matrix(model_data),
shap_approx = F)
```


```{r}
# Calculate data for SHAP plot
shap_long_1 = shap.prep(shap = shap_result_1,
      X_train =  as.matrix(model_data),
      top_n = 8)
```


```{r}
# Generate SHAP plot
plot.shap.summary(data_long = shap_long_1)
```


```{r}
set.seed(123456) # Set seed for reproducability
# Create cross-validation index
cv_ind <- sample(1:5, nrow(fg_data), replace = TRUE )
pred_store <- rep(NA, nrow(fg_data))
# Create accuracy store
cv_acc <- rep(NA, 5)
for(i in 1:5){ # For 1 to 5
  cv_train <- fg_data[cv_ind != i ,c(3:69)] # Create training data
  cv_test <- fg_data[cv_ind == i,  c(3:69)] # Create test data

  
  dtrain <- xgb.DMatrix(data = as.matrix(model_data[cv_ind != i ,]), label = kick_made[cv_ind != i])
  dtest <- xgb.DMatrix(data = as.matrix(model_data[cv_ind == i ,]), label = kick_made[cv_ind == i])
  
  
  xgb_1 <- xgboost(dtrain, 
                 objective = "binary:logistic", 
                 nrounds = 100)
  pred_store[cv_ind == i] <- predict(xgb_1, dtest) # Create test data predictions

}

# Print cross validated accuracy scores
cv_acc


fg_data$prob <- pred_store
fg_data$kick_made <- kick_made

```


```{r}
fg_data$diff <- fg_data$kick_made - fg_data$prob
```


```{r}
# Filter for clutch kicks and include kicker name
kicker_clutch_stats <- fg_data %>%
  filter(clutch == TRUE) %>%
  group_by(posteam, kicker_player_name) %>%
  summarize(
    total_clutch_kicks = n(),
    clutch_makes = sum(field_goal_result == "made"),
    clutch_fg_pct = mean(field_goal_result == "made"),
    total_wpa = sum(wpa, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  mutate(
    clutch_score = clutch_fg_pct * 0.5 + (total_wpa / total_clutch_kicks) * 0.5
  ) %>%
  arrange(desc(clutch_score))

# Add ranks
kicker_clutch_stats <- kicker_clutch_stats %>%
  mutate(rank = row_number())

# View results
print(kicker_clutch_stats)

```

```{r}
# Summarize expected vs actual FG makes by kicker
kicker_performance <- fg_data %>%
  group_by(kicker_player_name) %>%
  summarize(
    total_kicks = n(),
    actual_makes = sum(kick_made),
    expected_makes = sum(prob),
    diff = actual_makes - expected_makes
  ) %>%
  arrange(desc(diff))

# View the performance table
print(kicker_performance)

```

```{r}
library(ggplot2)

# Create a bar plot for expected vs actual makes
ggplot(kicker_performance, aes(x = reorder(kicker_player_name, diff), y = diff, fill = diff > 0)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  coord_flip() +  # Flip for better readability
  labs(
    title = "Expected vs Actual Field Goals Made by Kicker",
    x = "Kicker",
    y = "Actual Makes - Expected Makes"
  ) +
  scale_fill_manual(values = c("red", "green")) +  # Red for underperformance, green for overperformance
  theme_minimal()

```

